An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-29 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added updated all.bib with all old publications and use it in the text, added parts of PISA description (not done yet)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \cite{CAR08}. The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs. 
%
When solving such a problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions are not affected. In this case, one may wish to distribute the risk in an equitable way over the population and the environment. Several studies consider this additional minimization of the equity risk, but most of them consist of a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{AKG02, CAR08}).
% 
% 
% and can be classified into the \textit{resolution-equity-based methods} and the \textit{model-equity-based methods}. In resolution-equity-based methods, the equity constraints are taken into account only indirectly via a dissimilarity index between paths which is to be maximized \cite{JOH92, LOM93, KUB97, AKG02}. Model-equity-based methods, on the other hand, formalize the risk equity as constraints directly within the model. In \cite{GOP90a, GOP90b}, for example, the authors propose an equity shortest path model that minimizes the total risk of travel while the difference between the risks imposed on any two arbitrary zones does not exceed a given threshold.
%
A more realistic \emph{multi}-commodity flow model was proposed in \citep{CAR08} where each commodity is considered as one hazmat type. The objective function is formulated as the sum of the economical cost and the cost related to the consequences of an incident for each material. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs which imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.

The majority of all hazmat routing studies deal with a single-objective scenario although the problem itself is multiobjective in nature and it is important to study the trade-offs among the objectives. Evolutionary Multiobjective Optimization (EMO) algorithms are able to compute a set of solutions showing these trade-offs within a single algorithm run which is the reason why we propose to use them for the problem of hazmat routing in this study (Sec.~\ref{S_MEO}). Before, we formalize the routing of hazmat problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{CAR08} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).

% see for example \cite{CAR08, cgr2007, gkbk1990, za2004}, but all problem formulations include the equity risk as constraints and not as an additional objective. \citet{gian1998} on the other hand considers the equity risk as separate objective but uses a weighted sum to aggregate his four objectives into a single-objective problem. The main reason why, to the best of our knowledge, no study considered a fully multiobjective model with the equity risk as objective might be the resulting non-linearity of the problem which make classical optimization algorithms such as branch-and-cut \TODON{equity is an objective in \cite{{gian1998}} ... We solve the broblem with a branch and price algorithm because the model is a multicommodity flow model, and this model is the more adapted one for this problem because it permit to manage simultaneously many hazmat types. The min-max formulation of the objective is easily linearized. I give you the major axes of the work: (1) Solve the presented problem (optimize cost, risk and equity) (2) what are models and resolution methods proposed in the litterature (3) what is the model retained and why (in our case, the multicommodity model) (4) how to handle equity with this model ? (5) how to solve it ? (branch and price is adapted for these models, but can not hadle many objectives $->$ evolutionnary algorithms) } infeasible and led us to our choice of proposing an evolutionary multiobjective algorithm as heuristic here.

@@

%In \cite{CAR08} was proposed a multi-commodity flow model for this problem, where each commodity is considered as one hazmat type. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs, this imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc. The major drawback of this model is that it can lead to the saturation of the arcs of shortest paths and, therefore, does not guarantee the equity target.

@@

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}.

@@

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bianco09}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bianco09} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.

@@

%We concentrated our analysis on a real-world case study \citep{bianco09}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bianco09}. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \cite{cd2008a}. The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs. 
%
When solving such a problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions are not affected. In this case, one may wish to distribute the risk in an equitable way over the population and the environment. Several studies consider this additional minimization of the equity risk, but most of them consist of a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a, cd2008a}).
% 
% 
% and can be classified into the \textit{resolution-equity-based methods} and the \textit{model-equity-based methods}. In resolution-equity-based methods, the equity constraints are taken into account only indirectly via a dissimilarity index between paths which is to be maximized \cite{JOH92, LOM93, KUB97, aeb2000a}. Model-equity-based methods, on the other hand, formalize the risk equity as constraints directly within the model. In \cite{GOP90a, GOP90b}, for example, the authors propose an equity shortest path model that minimizes the total risk of travel while the difference between the risks imposed on any two arbitrary zones does not exceed a given threshold.
%
A more realistic \emph{multi}-commodity flow model was proposed in \citep{cd2008a} where each commodity is considered as one hazmat type. The objective function is formulated as the sum of the economical cost and the cost related to the consequences of an incident for each material. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs which imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc.

The majority of all hazmat routing studies deal with a single-objective scenario although the problem itself is multiobjective in nature and it is important to study the trade-offs among the objectives. Evolutionary Multiobjective Optimization (EMO) algorithms are able to compute a set of solutions showing these trade-offs within a single algorithm run which is the reason why we propose to use them for the problem of hazmat routing in this study (Sec.~\ref{S_MEO}). Before, we formalize the routing of hazmat problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}).

% see for example \cite{cd2008a, cgr2007, gkbk1990, za2004}, but all problem formulations include the equity risk as constraints and not as an additional objective. \citet{gian1998} on the other hand considers the equity risk as separate objective but uses a weighted sum to aggregate his four objectives into a single-objective problem. The main reason why, to the best of our knowledge, no study considered a fully multiobjective model with the equity risk as objective might be the resulting non-linearity of the problem which make classical optimization algorithms such as branch-and-cut \TODON{equity is an objective in \cite{{gian1998}} ... We solve the broblem with a branch and price algorithm because the model is a multicommodity flow model, and this model is the more adapted one for this problem because it permit to manage simultaneously many hazmat types. The min-max formulation of the objective is easily linearized. I give you the major axes of the work: (1) Solve the presented problem (optimize cost, risk and equity) (2) what are models and resolution methods proposed in the litterature (3) what is the model retained and why (in our case, the multicommodity model) (4) how to handle equity with this model ? (5) how to solve it ? (branch and price is adapted for these models, but can not hadle many objectives $->$ evolutionnary algorithms) } infeasible and led us to our choice of proposing an evolutionary multiobjective algorithm as heuristic here.

@@

%In \cite{cd2008a} was proposed a multi-commodity flow model for this problem, where each commodity is considered as one hazmat type. To deal with risk equity, the costs are defined as functions of the flow traversing the arcs, this imposes an increase of the arc's cost and risk when the number of vehicles transporting a given material increases on the arc. The major drawback of this model is that it can lead to the saturation of the arcs of shortest paths and, therefore, does not guarantee the equity target.

@@

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{}, and W-HypE \citep{} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}

$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.

 The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to

\TODO{describe how PISA works}

@@

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.

@@

%We concentrated our analysis on a real-world case study \citep{bcg2009a}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bcg2009a}. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.
